|

### 2021-04-25 index.Rmd
final pushes

--- 
title: "Bayesian Inference for Species Distribution Modelling with Gaussian Processes"
subtitle: "Isaac William Caruso"
author: |
  | Amherst College
  | Department of Computer Science
date: "`r Sys.Date()`"
output:
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
description: "Bayesian Inference for Species Distribution Modelling with Gaussian Processes"
documentclass: book
link-citations: yes
bibliography:
  - references.bib
  - packages.bib
site: bookdown::bookdown_site
biblio-style: apalike
fontsize: 12pt
linestretch: 1.7
geometry: "left=1in, right=1in, top=1in, bottom=1in"
---
```{r setup, include = FALSE}
library(tidyverse)
library(mosaic)
library(mdsr)
library(knitr)
library(kableExtra)
library(latex2exp)
library(extrafont)
library(rstan)
library(mdsr)
font_import(pattern = "Latin Modern Roman*") 
loadfonts()
setwd("~/bayes-gp-sdm-manuscript/")
```
# Probability and Bayes’ theorem

This chapter provides an introduction to the probability concepts necessary to understand Bayesian inference. 
Simply put, Bayesian inference is a statistical technique for estimating a quantity of interest upon the observation of data, while explicitly incorporating prior knowledge or belief about that quantity of interest.
Before embarking on an exposition of Bayesian statistics, we must first gain a basic understanding of a few key elements of probability theory—random variables, cumulative distribution functions, probability functions/ distributions, expected value, and conditional probability.
This chapter then concludes by introducing Bayes' theorem and Bayesian inference, and demonstrating the steps for performing a simple Bayesian update.

@@

Imagine for a moment you are tossing a fair coin. 
There are many experiments you could perform by tossing a coin, but let us consider our quantity of interest to be the fraction of times our coin lands on a tail. 
It is clear that the number of tails, the outcome of our experiment, is dependent on the eventual realization of some random process.
A random variable $X$ is a variable whose value is dependent on the outcome(s) of a stochastic phenomenon. 
The realized value of $X$ is denoted as $x$.

@@

In the previous example, we represented the probability of various outcomes of a coin toss experiment in a tabular format. 
Another way to represent this set of probabilities is as a *cumulative distribution function*.

```{definition, cdf, name="Cumulative distribution function 'CDF'"}
*The cumulative distribution function is defined as a function where $F_{X}\in[0, 1]$:*
$$F_{X}(x) \doteq P(X≤x)$$

@@

(ref:cdffig) CDF for tossing a coin twice [@Wasserman2004]

@@

Here, for every value of $x$—the number of tails in two coin tosses—the probability that $X$ is equal to or less than this value is represented.
In this discrete example, we see that our CDF is represented by several non-decreasing discrete lines defined for all $x$. 
In the example of a continuous random variable, this function is a left-continuous, non-decreasing distribution also defined for all $x$.

@@

The probability mass function and probability density function allow us to express probabilities of events over a sample space and find their use with discrete and continuous random variables respectively.

In the discrete setting, the probability mass function for a random variable $X$ yields the probability that $X$ takes a value for every possible value that $X$ can take.

```{definition, pmf, name="Probability mass function 'PMF'"}
*The probability function for a discrete random variable $X$—the probability mass function for $X$—is defined as a function*
$$f_{X}(x) \doteq P(X=x)$$

@@

Here, the PMF has a few key attributes. 
Namely $P(X=x)>0$ for every $x$ in the sample space $S_{X}$ of $X$, and $\sum_{x \in S_{X}}f_{X}(x)=1$. 
With these features in mind, the probability mass function of $X$ follows logically from the cumulative distribution function of $X$ insofar as the CDF is the sum of the PMF for all $x_{i}≤x$, i.e., $$F_{X}(x) \doteq P(X≤x)=\sum_{x_i≤x}f_{X}(x_{i}).$$

In the case where the random variable $X$ is continuous, its PDF is defined as follows.

```{definition, pdf, name="Probability density function 'PDF'"}
*The probability function for a continuous random variable $X$—the probability density function for $X$—is defined as a function $f(x)$ where $a$ and $b$ are two real numbers such that $a ≤ b$, so*
$$P(a<X<b) \doteq \int_{a}^{b}f_X(x)dx.$$

@@

One of the final core statistical concepts necessary to approach Bayesian statistics on sure footing is the idea of expectation or expected value.

@@

To return to a familiar example, consider a random variable $X$ to represent the number of tails in 6 coin tosses. Figure \@ref(fig:binom) depicts the PMF for $X$ using the *binomial distribution* $B(6,0.5)$ which represents the probability of observing a specific number of successes in a success-failure experiment.

@@

Conditional probability provides a way to model the probability that an event occurs, given that another event is known to have occurred.
Conditional probability, as presented in Definition \@ref(def:condprob), requires only that the event assumed to have occurred, i.e., the event we are *conditioning on*, has a nonzero probability of occurring.

@@

Conditional probability asserts that the probability of event A occurring given that event B occurs is equivalent to the probability of both A and B occurring (denoted as $A \cap B$) divided by the probability that B occurs. $P(A|B)$ is not generally equal to $P(B|A)$ 
For example, the probability that I am swimming given that I am in the water is clearly not the same as the probability that I am in the water given that I am swimming.
Example \@ref(exm:condprobe) explains how to use conditional probability to calculate the probability of rolling a 2 on a 6-sided dice, given I know the outcome of the roll is less than 4.

@@

Consider a student, Alice, who was exposed to someone with COVID-19. 
Being a responsible person, Alice decides she should get tested. 
She receives a test and the accompanying information sheet states the test is 85% accurate, meaning that 85% of the time it gives positive results to recipients who are actually positive. The sheet also says that the test yields a false positive 30% of the time, meaning that if Alice is actually negative she will still receive a positive test 30% of the time. 
The following day Alice receives a positive test.
As a student of probability, Alice recognizes that the 85% accuracy statistic only means the conditional probability that she receives a positive test given she is COVID positive ($P(+ test|covid+)$) is 85%.
However, Alice is really interested in the conditional probability that she is positive given she just tested positive, $P(covid+|test+)$.
As stated in the previous section's discussion of conditional probability, $P(A|B) \neq P(B|A)$.
Bayes' theorem, which follows intuitively from the theorem of conditional probability, provides this answer for Alice.
We can rewrite Definition \@ref(def:condprob) as 
$$P(A \cap B)=P(A|B)P(B).$$
Furthermore, it can also be stated that
$$P(B \cap A)=P(B|A)P(A)$$
Clearly, $P(A \cap B)=P(B \cap A)$ as both terms can be used interchangeably to represent the intersection of two sets A and B.
This equivalency means that we can rewrite these equations as
$$P(B|A)P(A)=P(A|B)P(B)$$
Dividing both sides of this equivalency by $P(B)$ yields Bayes' theorem, as formalized in Definition \@ref(def:bayestheorem) [@Junker2003].

```{definition, bayestheorem, name="Bayes' Theorem"}
*Assuming $P(B)>0$ and $P(A)>0$,*
$$P(B|A)=\frac{P(A|B)P(B)}{P(A)}$$
```

Additionally, the *law of total probability*, Definition \@ref(def:totprob), can be used to compute $P(A)$ for a discrete sample space $S_{B}$ [@Wasserman2004].

@@

In the previous example, we used Bayes' theorem to evaluate the probability that Alice is COVID positive given she received a positive test. 
Bayes' theorem provided a way for Alice to apply her prior beliefs about the likelihood of actually being positive to the data she observed---a positive test. 
*Learning*, or *statistical inference*, asks a generalized version of Alice's question. 
Given some data $X_0,...,X_i \sim D$, how can we infer the distribution $D$ that generated the data $X_0,...,X_i$ [@Wasserman2004 \text{Sec} 6.1]?
Additionally, as distributions are functions of parameters $\theta_0, ..., \theta_j$, estimating these parameters given the data is one avenue for inferring $D$.
*Bayesian inference* applies Bayes' theorem to the task of statistical inference in an intuitive manner, which by definition maintains the inherent uncertainty in inference.
Contrary to the frequentist perspective, a Bayesian approach to inference expresses degrees of belief by producing posterior probability distributions for the parameters $\theta_0, ..., \theta_j$ of the distribution [@Wasserman2004 \text{Sec} 11.1]. 
Point estimates and confidence/posterior intervals can then be computed with a post hoc analysis of the posterior distribution.
Furthermore, a Bayesian approach to inference explicitly includes previous domain-specific knowledge (or lack thereof) in the learning process.
Performing Bayesian inference to compute a posterior distribution $P(\theta|X)$ for a parameter $\theta$ first requires choosing a *prior probability distribution* $P(\theta)$ with a PDF $f(\theta)$ which expresses our prior beliefs about $\theta$ before observing any data. 
In the previous example, this was Alice's previous belief about the probability of contracting COVID from any given exposure.
Expressing $P(\theta)$ as a distribution allows for the inclusion of the inherent uncertainty of Alice's preconceptions in the modeling schema and thus the computation of the posterior.
Secondly, the likelihood function $l(X|\theta)$ must be specified to express our understanding of the probability of the data given a particular value of $\theta$.
In the case of Alice's inquiry, a binomial likelihood function would be appropriate as it represents the likelihood of observing a given number of successes ($+$tests) in $0 \lor 1$ ($+ \lor -$) trials.
The final step in performing Bayesian inference is repeatedly applying Bayes' theorem to update the probability distribution of $\theta$ for each observation $X_n$ in the data $n \in (0,i)$, resulting in a posterior distribution $f(\theta|X_0,...,X_i)$.
This is done by iteratively evaluating Bayes' theorem, where the posterior for the $n-1^{th}$ iteration is the prior for the $n^{th}$ iteration.
Bayesian inference relies on Bayes' theorem for continuous variables, which utilizes density functions as presented formally in Definition \@ref(def:bayesprob).

@@

While Bayesian inference is an appealing schema, learning the parameters of real-world models using this technique requires the use of efficient algorithms to be computationally feasible.
As previously mentioned, depending on the system we may be interested in evaluating an arbitrary number of parameters $\theta_0, ..., \theta_j$ given the data, and each parameter may potentially take on an infinite number of values.
Even in the case where we only consider a small number of possible values $a$ for every $\theta_i \in (0,j)$, the evaluation of the posterior to perform a single update rapidly becomes a computationally intractable task.
This is evidenced by the runtime of the *grid approximation* or *direct discrete approximation*, which is proportional to $O(a^{j})$, where $a$ controls the granularity of the approximation [@Gelman2013 \text{Sec} 10.1].
To perform Bayesian inference on complex systems, fast algorithms are necessary to circumnavigate this curse of dimensionality.

@@

When performing Bayesian inference, fast algorithms are necessary for updating the probability distributions of parameters of interest in a computationally feasible manner.
In modeling scenarios with multiple parameters forming a multidimensional parameter space, evaluating parameters at even a relatively small number of possible values rapidly becomes intractable, as a product of the curse of dimensionality.
This problem is further exaggerated in the evaluation of hierarchical models, which are of particular interest in many application domains such as biology, economics, chemistry, and physics.
Here we discuss Markov Chain Monte Carlo (MCMC) based sampling algorithms, which allow for efficiently sampling from approximations of probability distributions.
Markov Chain Monte Carlo (MCMC) provides a methodology for sampling from the posterior curve to perform a Bayesian update. 
Simply put, if we make some sequence of draws $\theta_1,...,\theta_j$ from the posterior $p(\theta|X)$, then plotting $\theta_1,...,\theta_j$ as a histogram will provide an approximation of $p(\theta|X)$ [@Wasserman2004 \text{Sec} 11.4].
Due to the law of large numbers, this approximation can be used in place of the posterior in post-hoc analysis.
We first provide a theoretical background into Monte Carlo approximation and Markov Chains, followed by a discussion of *Metropolis-Hastings*, an algorithm for performing MCMC.

@@

Monte Carlo approximation provides a convenient method for estimating quantities of interest.
This schema will allow for drawing samples from arbitrarily complex probability distributions.
The basic example is Monte Carlo *integration* [@Wasserman2004 \text{Sec} 24.2]. 
If we want to evaluate an integral for some function $f(x)$, where 
$$I=\int_{a}^{b}f(x)dx \textrm{ ,}$$
we can approximate $I$ using Monte Carlo approximation. $f(x)$ can be alternatively expressed as two functions, $h(x)$ and $w(x)$, where $h(x)=\frac{1}{b-a}$ and $w(x)=f(x)(b-a)$ as follows.
$$I=\int_{a}^{b}f(x)dx=\int_{a}^{b}w(x)h(x)dx$$
Conveniently, $h$ is the pdf of a uniform r.v. $X$ over $[a,b]$, which means that $I$ can be rewritten in terms of expectation as
$$I=E_{f}(w(X)) \textrm{ .}$$
In conjunction with the law of large numbers, this means that if we generate a sequence of random variables from a uniform distribution $X_{0},...,X_{N}\sim unif(a,b)$ then the standard Monte Carlo integration method asserts

$$\hat{I}\equiv\frac{1}{N}\sum_{i=0}^{N}w(X_{i})\rightarrow E(w(X))=I \textrm{ .}$$

@@

```{example montecarl}
Suppose we have two binomially distributed r.vs---$X \sim Binom(n,p_1)$, $Y \sim Binom(m,p_2)$---and want to evaluate $\delta = p_2 - p_1$. 
This problem can be analyzed using Bayesian inference on a beta-binomial model but would require evaluating complex integrals over the joint posterior of $X$ and $Y$ to obtain the density.
Monte Carlo Integration allows for estimating an integral using repeated sampling from the two independent posterior densities---$P_1$ and $P_2$--- and plotting samples using a histogram provides a qualitative estimate for the parameter of interest, $\delta$ [@Wasserman2004].
In simplest terms, the beta prior distribution $Beta(\alpha, \beta)$ (the proper conjugate prior for the binomial likelihood) could be a flat prior $\alpha = \beta = 1$, on the parameters $p_1$ and $p_2$ [@Bloom2014].
$f(p_1,p_2|X,Y)=f(p_1|X)f(p_2|Y)$, because the two posteriors $f(p_1|X)$ and $f(p_2|Y)$ are conditionally independent given the data.
Thus the resultant posteriors are beta distributions, as follows [@Bloom2014].
$$D_1 = Beta(X+1,n-X+1)$$
$$D_2 = Beta(Y+1,m-Y+1)$$
Monte Carlo integration simulates the quantity $\delta$ by drawing $(P_1^{(1)},P_2^{(2)}),...,(P_1^{(N)},P_2^{(N)})$ from the independent beta posteriors $D_1$ and $D_2$ for $i=1,...,N$ assuming N is sufficiently large, every $\delta^{(i)}=P_2^{(i)} - P_1^{(i)}$.
Accordingly, $\delta$ is approximated using the previously established equivalency, $$\delta \approx \frac{1}{N} \sum_i^N \delta^{(i)} = \frac{1}{N} \sum_i^N (P_2^{(i)} - P_1^{(i)})$$ [@Wasserman2004].
To provide a concrete example, if the number of trials is 20, $n=m=20$, $X = 11$, and $Y = 4$ Sampling 10000 times from the in the previously described manner yields the histogram presented in Figure \@ref(fig:mon) 
```
```{r mon, fig.height=3, fig.align='center', echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Histogram displaying posterior of delta from Monte Carlo integration"}
p1 <- rbeta(10000, 12, 10)
p2 <- rbeta(10000, 5, 17)
delta <- p2 - p1
delta_df <- data.frame(delta = delta)
hist <- ggplot(delta_df, aes(x=delta)) + 
  geom_histogram(colour="black", fill="#856cb0") +
  theme_mdsr() +
  theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.title = element_blank())
  #theme(text = element_text(size=12, 
  #                          family = "CMU Serif",
   #                         face = "plain",
   #                         colour = "black",
    #                        hjust = 0.5,
     #                       vjust = 0.5,
      #                      angle = 0,
       #                     lineheight = 0.9))

@@

## Markov Chains

A Markov chain is a stochastic model expressing a sequence of possible states in which the probability of each state, $X_{i}$, depends only on the value attained in the previous state, $X_{i-1}$, for all $i \geq 0$.
A Markov chain has several properties which are essential for its application in Bayesian statistics. 
Definition \@ref(def:markov) formalizes the requirements for a Markov chain [@Wasserman2004].

>>>>

--- 
title: |
  | Bayesian Inference with Stan 
  | for Hybrid Species Distribution Modeling
subtitle: |
  |
  | \vspace{+20truemm}
  | Isaac William Caruso
  | Advisor: Prof. Matteo Riondato
  | April 26, 2021
author: |
  | \vspace{+45truemm}
  | Submitted to the Department of Computer Science of Amherst College
  | in partial fulfillment of the requirements for the degree of
  | Bachelor of Arts with Distinction
output:
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
description: "Bayesian Inference with Stan for Hybrid Species Distribution Modeling"
documentclass: book
link-citations: yes
bibliography:
  - references.bib
  - packages.bib
site: bookdown::bookdown_site
biblio-style: apalike
fontsize: 12pt
linestretch: 1.7
geometry: "left=1in, right=1in, top=1in, bottom=1in"
classoption: twoside
indent: true
---
```{r setup, include = FALSE}
library(tidyverse)
library(mosaic)
library(mdsr)
library(knitr)
library(kableExtra)
library(latex2exp)
library(extrafont)
library(rstan)
library(brms)
library(bayesplot)
library(formatR)
font_import(pattern = "Latin Modern Roman*") 
loadfonts()
setwd("~/bayes-gp-sdm-manuscript/")
```

# Abstract {-}

Bayesian inference is a statistical technique for estimating a quantity of interest from data.
Increasingly, Bayesian inference is being applied to problems in a variety of fields of study as it allows for the explicit incorporation of *a priori*, *domain-specific* knowledge into the learning process.
While generally appealing for numerous reasons, it has until recently been an impractical tool for learning the parameters of complex, real-world systems.
In this work, we first provide the theoretical concepts necessary for appreciating Bayesian inference.
We then describe algorithms for performing Bayesian inference with computing machines in a computationally feasible manner, namely Metropolis-Hastings Markov Chain Monte Carlo (MCMC).
This is then followed by a discussion of Stan, a domain-specific programming language often used for MCMC and Bayesian inference.
Finally, we conclude this work with a case study from a recent publication in the field of biology, which applies Bayesian inference to the task of species distribution modeling.
Not only is this task extremely pertinent as it relates directly to climate change, but also the resulting model was sufficiently complex to test the full capabilities of the Stan language.

# Acknowledgments {-}

I would like to thank all of the Amherst College Computer Science Department---especially my thesis advisor Professor Matteo Riondato---for facilitating my academic growth over the past four years and in the endeavor of completing this thesis. 
Moreover, I thank Professor Lauren Buckley from the University of Washington Department of Biology for graciously providing her expertise and guidance in Biology and Functional Ecology.
I also am excited to acknowledge my parents for their continued support over the entirety of my educational career and my friends, who provided an extensive network of support during my time at Amherst.

# Introduction {-}

As data collection becomes increasingly commonplace, extracting meaning from large datasets is essential for drawing insights and predictions.
Understanding how a large number of factors influence a dataset is a topic of certain relevance for almost every scientific domain.
In the field of biology, as we expand upon in the fourth chapter, understanding where animal species occur in the wild and how these ranges are shifting over time due to environmental factors is of critical importance for conservation efforts to have the greatest possible effect.
Conservation biologists and ecologists have been studying these species' tolerances to the environment for quite some time, so we understand how different properties of the environment influence the occurrence of species we are interested in.
Additionally, we are not only interested in where these species have been observed, but also where they have *not* been observed.
The fields of statistics and machine learning provides the computational tools necessary for addressing this inquiry and many others.
Given existing data, we learn a *model* which incorporates information we already have and the data we have available to us to predict what a situation for which we do not yet have the data will look like. 
As we elaborate upon in the first chapter of this work, Bayesian approaches to learning allow us to incorporate *a priori* domain-specific knowledge in a manner that is both explicit and intuitive.
The second chapter discusses *Markov Chain Monte Carlo* (MCMC) and related algorithmic developments that make Bayesian inference feasible and scalable.
Chapter three presents a discussion of Stan, a software project aimed at making Bayesian inference fast and accessible. 
Since at the outset Bayesian inference may appear quie complex, we aim to provide a gradual introduction first to probability and the concepts necessary to comprehend Bayesian statistics and secondly to how performing Bayesian inference is possible using modern computational techniques.

# Probability primer

This chapter provides an introduction to the probability concepts necessary to understand Bayesian inference. 
Simply put, Bayesian inference is a statistical technique for estimating a quantity of interest upon the observation of data, while explicitly incorporating *prior knowledge or belief* about that quantity of interest.
Before embarking on an exposition of Bayesian statistics, we must first gain a basic understanding of a few key elements of probability theory—random variables, cumulative distribution functions, probability functions/ distributions, expected value, and conditional probability.
This chapter concludes by introducing Bayes' theorem and Bayesian inference.

@@

Imagine for a moment you are tossing a fair coin. 
There are many experiments you could perform by tossing a coin, but let us consider our quantity of interest to be the fraction of times our coin lands on a tail. 
It is clear that the number of tails, the outcome of our experiment, is dependent on the eventual realization of some random process.
A *random variable* $X$ is a variable whose value is dependent on the outcome(s) of a stochastic phenomenon. 
The realized value of $X$ is denoted as $x$.

@@

In the previous example, we represented the probability of various outcomes of a coin toss experiment in a tabular format. 
Another way to represent this set of probabilities is as a *cumulative distribution function* [@Wasserman2004 \text{Sec} 2.2].

```{definition, cdf, name="Cumulative distribution function 'CDF'"}
*The cumulative distribution function is defined as the function*
$$F_{X}(x) \doteq P(X≤x)$$

@@

(ref:cdffig) CDF for tossing a coin twice [@Wasserman2004 \text{Sec} 2.2]

@@

Here, for every value of $x$—the number of tails in two coin tosses—the probability that $X$ is equal to or less than this value is represented.
In this discrete example, we see that our CDF is represented by several non-decreasing discrete lines defined for all $x$. 
In the example of a continuous random variable, this function is a non-decreasing distribution also defined for all $x$.

@@

The *probability mass function* (PMF) and *probability density function* (PDF) allow us to express probabilities of events over a sample space and find their use with discrete and continuous random variables respectively.

In the discrete setting, the probability mass function for a random variable $X$ yields the probability that $X$ takes a value $x$ for every possible value that $X$ can take [@Wasserman2004 \text{Sec} 2.2].

```{definition, pmf, name="Probability mass function 'PMF'"}
*The probability function for a discrete random variable $X$—known as the *probability mass function* for $X$—is defined as the function*
$$f_{X}(x) \doteq P(X=x)$$

@@

Here, the PMF has a few key attributes. 
Namely $P(X=x)>0$ for every $x$ in the sample space $S_{X}$ of $X$, and $\sum_{x \in S_{X}}f_{X}(x)=1$. 
With these features in mind, the probability mass function of $X$ follows logically from the cumulative distribution function of $X$ insofar as the CDF is the sum of the PMF for all $x_{i}≤x$ [@Wasserman2004 \text{Sec} 2.2], i.e., $$F_{X}(x) \doteq P(X≤x)=\sum_{x_i≤x}f_{X}(x_{i}).$$

In the case where the random variable $X$ is continuous, its probability density function is defined as follows [@Wasserman2004 \text{Sec} 2.2].

```{definition, pdf, name="Probability density function 'PDF'"}
*The probability function for a continuous random variable $X$—known as the *probability density function* for $X$—is defined as a function $f(x)$ where $a$ and $b$ are two real numbers such that $a ≤ b$, so*
$$P(a<X<b) \doteq \int_{a}^{b}f_X(x)dx.$$

@@

One of the final core probability concepts necessary to approach Bayesian statistics on sure footing is the idea of expectation or expected value [@Wasserman2004 \text{Sec} 3.1].

@@

To return to a familiar example, consider a random variable $X$ to represent the number of tails in 6 coin tosses. Figure \@ref(fig:binom) depicts the PMF for $X$ using the *binomial distribution* $B(6,0.5)$ which represents the probability of observing a specific number of successes in a success-failure experiment, where the probability of success is 0.5.

@@

*Conditional probability* provides a way to model the probability that an event occurs, given that another event is known to have occurred.
Conditional probability, as presented in Definition \@ref(def:condprob), requires only that the event assumed to have occurred, i.e., the event we are *conditioning on* has a nonzero probability of occurring [@Wasserman2004 \text{Sec} 3.5].

@@

Conditional probability asserts that the probability of event A occurring given that event B occurs is equivalent to the probability of both A and B occurring (denoted as $A \cap B$) divided by the probability that B occurs. $P(A|B)$ is not generally equal to $P(B|A)$.
For example, the probability that I am swimming given that I am in the water is clearly not the same as the probability that I am in the water given that I am swimming.
Example \@ref(exm:condprobe) explains how to use conditional probability to calculate the probability of rolling a 2 on a 6-sided dice, given I know the outcome of the roll is less than 4.

@@

Consider a student, Alice, who was exposed to someone with COVID-19. 
Being a responsible citizen, Alice decides she should get tested. 
She receives a test and the accompanying information sheet states the test is 85% accurate, meaning that 85% of the time it gives positive results to recipients who are actually positive. The sheet also says that the test yields a false positive 30% of the time, meaning that if Alice is actually negative she will still receive a positive test 30% of the time. 
The following day Alice receives a positive test.
As a student of probability, Alice recognizes that the 85% accuracy statistic only means the conditional probability that she receives a positive test given she is COVID positive ($P(+ test|covid+)$) is 85%.
However, Alice is really interested in the conditional probability that she is positive given she just tested positive, $P(covid+|test+)$.
As stated in the previous section's discussion of conditional probability, $P(A|B) \neq P(B|A)$.
*Bayes' theorem*, which follows intuitively from the theorem of conditional probability, provides this answer for Alice.
We can rewrite Definition \@ref(def:condprob) as 
$$P(A \cap B)=P(A|B)P(B).$$
Furthermore, it can also be stated that
$$P(B \cap A)=P(B|A)P(A)$$
Clearly, $P(A \cap B)=P(B \cap A)$ as both notations can be used interchangeably to represent the intersection of two sets A and B.
This identity means that we can rewrite the previous equations as
$$P(B|A)P(A)=P(A|B)P(B)$$
Moreover, dividing both sides of this equivalency by $P(A)$ yields Bayes' theorem, as formalized in Definition \@ref(def:bayestheorem) [@Junker2003].

```{definition, bayestheorem, name="Bayes' Theorem"}
*Assuming $P(A)>0$,*
$$P(B|A)=\frac{P(A|B)P(B)}{P(A)}$$
```

The *law of total probability*, Defin-ition \@ref(def:totprob), can be used to compute $P(A)$ for a discrete sample space $S_{B}$ [@Wasserman2004 \text{Sec} 1.7].

@@

In the previous example, we used Bayes' theorem to evaluate the probability that Alice is COVID positive given she received a positive test. 
Bayes' theorem provided a way for Alice to apply her prior beliefs about the likelihood of actually being positive to the data she observed---a positive test. 
*Learning*, or *statistical inference*, asks a generalized version of Alice's question. 
Given some data $X_0,...,X_i \sim D$, how can we infer the distribution $D$ that generated the data $X_0,...,X_i$ [@Wasserman2004 \text{Sec} 6.1]?
Additionally, as distributions are functions of parameters $\theta_0, ..., \theta_j$, estimating these parameters given the data is one avenue for inferring $D$.
*Bayesian inference* applies Bayes' theorem to the task of statistical inference in an intuitive manner, which by definition maintains the inherent uncertainty in inference.
Contrary to the frequentist perspective, a Bayesian approach to inference expresses degrees of belief by producing *posterior probability distributions* for the parameters $\theta_0, ..., \theta_j$ of the distribution [@Wasserman2004 \text{Sec} 11.1]. 
Point estimates and confidence/posterior intervals can then be computed with a post hoc analysis of the posterior distribution.
Furthermore, a Bayesian approach to inference explicitly includes previous domain-specific knowledge (or lack thereof) in the learning process.
Performing Bayesian inference to compute a posterior distribution $P(\theta|X)$ for a parameter $\theta$ first requires choosing a *prior probability distribution* $P(\theta)$ with a PDF $f(\theta)$ which expresses our prior beliefs about $\theta$ before observing any data. 
In the previous example, this was Alice's previous belief about the probability of contracting COVID from any given exposure.
Expressing $P(\theta)$ as a distribution allows for the inclusion of the inherent uncertainty of Alice's preconceptions in the modeling schema and thus the computation of the posterior.
Secondly, the likelihood function $l(X|\theta)$ must be specified to express our understanding of the probability of the data given a particular value of $\theta$.
In the case of Alice's inquiry, a binomial likelihood function would be appropriate as it represents the likelihood of observing a given number of successes ($+$tests) in $0 \lor 1$ ($+ \lor -$) trials.
The final step in performing Bayesian inference is repeatedly applying Bayes' theorem to update the probability distribution of $\theta$ for each observation $X_n$ in the data $n \in (0,i)$, resulting in a posterior distribution $f(\theta|X_0,...,X_i)$.
This is done by iteratively evaluating Bayes' theorem, where the posterior for the $n-1^{th}$ iteration is the prior for the $n^{th}$ iteration.
Bayesian inference relies on Bayes' theorem for continuous variables, which utilizes density functions as presented formally in Definition \@ref(def:bayesprob) [@Junker2003].

@@

While Bayesian inference is an appealing schema, learning the parameters of real-world models using this technique requires the use of efficient algorithms to be computationally feasible.
As previously mentioned, depending on the system we may be interested in evaluating an arbitrary number of parameters $\theta_0, ..., \theta_j$ given the data, and each parameter may potentially take on an infinite number of values.
Even in the case where we only consider a small number of possible values $a$ for every $\theta_i \in (0,j)$, the evaluation of the posterior to perform a single update rapidly becomes a computationally intractable task.
This is evidenced by the runtime of the *grid approximation* or *direct discrete approximation*, which is proportional to $O(a^{j})$, where $a$ controls the granularity of the approximation [@Gelman2013 \text{Sec} 10.1].
To perform Bayesian inference on complex systems, fast algorithms are necessary to circumvent this curse of dimensionality.

@@

When performing Bayesian inference, fast algorithms are necessary for computing the posterior probability distributions of parameters of interest in a computationally feasible manner.
In modeling scenarios with multiple parameters forming a multidimensional parameter space, evaluating parameters at even a relatively small number of possible values rapidly becomes intractable, as a product of the curse of dimensionality.
This problem is further exaggerated in the evaluation of hierarchical models, which are of particular interest in many application domains such as biology, economics, chemistry, and physics.
Here we discuss Markov Chain Monte Carlo (MCMC) based sampling algorithms, which allow for efficiently sampling from approximations of probability distributions.
MCMC provides a methodology for sampling from the posterior distribution to perform a Bayesian update. 
Simply put, if we make some sequence of draws $\theta_1,...,\theta_j$ from the posterior $p(\theta|X)$, then plotting $\theta_1,...,\theta_j$ as a histogram will provide an approximation of $p(\theta|X)$ [@Wasserman2004 \text{Sec} 11.4].
Due to the law of large numbers, this approximation can be used in place of the posterior in post hoc analysis to derive point estimates and confidence intervals.
We first provide a theoretical background into Monte Carlo approximation and Markov chains, followed by a discussion of *Metropolis-Hastings*, an algorithm for performing MCMC.

@@

Monte Carlo approximation provides a convenient method for estimating quantities of interest.
This schema will allow for drawing samples from arbitrarily complex probability distributions.
The basic example is *Monte Carlo integration* [@Wasserman2004 \text{Sec} 24.2]. 
If we want to evaluate an integral for some function $f(x)$, where 
$$I=\int_{a}^{b}f(x)dx \textrm{ ,}$$
we can approximate $I$ using Monte Carlo approximation. $f(x)$ can be alternatively expressed as two functions, $h(x)$ and $w(x)$, where $h(x)=\frac{1}{b-a}$ and $w(x)=f(x)(b-a)$ as follows.
$$I=\int_{a}^{b}f(x)dx=\int_{a}^{b}w(x)h(x)dx$$
Conveniently, $h$ is the pdf of a uniform r.v. $X$ over $[a,b]$, which means that $I$ can be rewritten in terms of expectation as
$$I=E_{f}(w(X)) \textrm{ .}$$
In conjunction with the law of large numbers, this means that if we generate a sequence of random variables from a uniform distribution $X_{0},...,X_{N}\sim unif(a,b)$ then the standard Monte Carlo integration method asserts

$$\hat{I}=\frac{1}{N}\sum_{i=0}^{N}w(X_{i})\rightarrow E(w(X))=I \textrm{ .}$$

@@

```{example montecarl}
Suppose we have two binomially distributed r.vs---$X \sim Binom(n,p_1)$, $Y \sim Binom(m,p_2)$---and want to evaluate $\delta = p_2 - p_1$. 
This problem can be analyzed using Bayesian inference on a beta-binomial model but would require evaluating complex integrals over the joint posterior of $X$ and $Y$ to obtain the density.
Monte Carlo Integration allows for estimating an integral using repeated sampling from the two independent posterior densities---$P_1$ and $P_2$--- and plotting samples using a histogram provides a qualitative estimate for the parameter of interest, $\delta$ [@Wasserman2004 \text{Sec} 24.2].
In simplest terms, the prior distribution could be a flat prior on the parameters $p_1$ and $p_2$ [@Wasserman2004 \text{Sec} 24.2, @Bloom2014].
$f(p_1,p_2|X,Y)=f(p_1|X)f(p_2|Y)$, because the two posteriors $f(p_1|X)$ and $f(p_2|Y)$ are conditionally independent given the data.
Thus the resultant posteriors are beta distributions, as follows [@Bloom2014].
$$D_1 = Beta(X+1,n-X+1)$$
$$D_2 = Beta(Y+1,m-Y+1)$$
Monte Carlo integration simulates the quantity $\delta$ by drawing $(P_1^{(1)},P_2^{(2)}),...,(P_1^{(N)},P_2^{(N)})$ from the independent beta posteriors $D_1$ and $D_2$ for $i=1,...,N$, and every $\delta^{(i)}=P_2^{(i)} - P_1^{(i)}$.
Assuming N is sufficiently large, $\delta$ is approximated using the previously established equivalency, $$\delta \approx \frac{1}{N} \sum_i^N \delta^{(i)} = \frac{1}{N} \sum_i^N (P_2^{(i)} - P_1^{(i)})$$ [@Wasserman2004 \text{Sec} 24.2].
To provide a concrete example, if the number of trials is 20, $n=m=20$, $X = 11$, and $Y = 4$, sampling 10000 times from the posterior in the previously described manner yields the histogram presented in Figure \@ref(fig:mon).
We can then use the histogram to give a point estimate of, e.g., -0.32 to $\delta$.
```
```{r mon, fig.height=3, fig.align='center', echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Histogram displaying posterior of delta from Monte Carlo integration"}
p1 <- rbeta(10000, 12, 10)
p2 <- rbeta(10000, 5, 17)
delta <- p2 - p1
delta_df <- data.frame(delta = delta)
hist <- ggplot(delta_df, aes(x=delta)) + 
  geom_histogram(colour="black", fill="#856cb0") +
  theme_mdsr() +
  theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.title = element_blank())
  #theme(text = element_text(size=12, 
  #                          family = "CMU Serif",
   #                         face = "plain",
   #                         colour = "black",
    #                        hjust = 0.5,
     #                       vjust = 0.5,
      #                      angle = 0,
       #                     lineheight = 0.9))

@@

## Markov chains

A Markov chain is a stochastic model expressing a sequence of possible states in which the probability of each state, $X_{i}$, depends only on the value attained in the previous state, $X_{i-1}$, for all $i \geq 0$.
A Markov chain has several properties which are essential for its application in Bayesian statistics. 
Definition \@ref(def:markov) formalizes the requirements for a Markov chain [@Wasserman2004 \text{Sec} 23.2].
